### 2024-08-24 paper.tex
move determinant calculation into linalg section
fixup notation in proof of jordan basis, up to the end of 'proof of (1)'

@@

\begin{lemma}
    Let \(X \in \gl_m\), and let 
    \[Y = \begin{pmatrix}
        y_{11} & y_{12} & y_{13} & \cdots & y_{1,n-1} & y_{1n} \\
        d_1 & y_{22} & y_{23} & \cdots & y_{2,n-1} & y_{2n} \\
            & d_2   & y_{33} & \cdots & y_{3,n-1} & y_{3n} \\
            & & \ddots & \cdots & \vdots  & \vdots \\
            & & & d_{n - 2} & y_{n-1,n-1} & y_{n - 1,n}\\
            & & & & d_{n - 1} & y_{nn}
    \end{pmatrix} \in \gl_n.\]
    For any \(a, b \in \mathbb{C}^m\),
    \[\det 
    \begin{pNiceArray}{ccc|ccc}
        & & & & & \vert \\
        & X & & & & b    \\
        & & & & & \vert \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y & \\
       & & &  \\
       \end{pNiceArray} = \]
       \[\det X \det Y + \left(\prod_i d_i\right) \det \begin{pNiceArray}{ccc|c}
            & & & \vert \\
            & X & & b     \\
            & & & \vert \\
            \hline 
        \text{---} & a & \text{---} & 0
       \end{pNiceArray}\]
\end{lemma}
\begin{proof}
    By induction on \(n\).
    In the case \(n = 1\), expanding along the last row (taking the usual interpretation of the empty product) gives the desired result.
    \par Now suppose \(n > 1\).
    Expanding along the last row, we get 
    \[d_{n - 1} \det \begin{pNiceArray}{ccc|ccc}
        & & & & & \vert \\
        & X & & & & b    \\
        & & & & & \vert \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y_{n,n-1} & \\
       & & &  \\
       \end{pNiceArray} -
    y_{nn} \det \begin{pNiceArray}{ccc|ccc}
        & & & & & \\
        & X & & & &   \\
        & & & & & \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y_{n,n} & \\
       & & &  \\
       \end{pNiceArray}.\]
    Using our inductive hypothesis for the first determiniant, and using that \(\det \begin{pNiceArray}{c|c}
        A_{11} & 0 \\
        A_{21} & A_{22}
    \end{pNiceArray} = \det A_{11} \det A_{22}\) for the second, the expression becomes
    \[d_{n - 1} \left(\det X \det Y_{n,n-1} + \left(\prod_{i \leq n - 2} d_i\right) \det \begin{pNiceArray}{ccc|c}
        & & & \vert \\
        & X & & b     \\
        & & & \vert \\
        \hline 
    \text{---} & a & \text{---} & 0
   \end{pNiceArray}\right)
   - y_{nn} \det X \det Y_{nn} = \]
   \[(d_{n - 1} Y_{n,n-1} - y_{nn} \det Y_{nn})\det X + \left(\prod_i d_i\right) \det \begin{pNiceArray}{ccc|c}
        & & & \vert \\
        & X & & b     \\
        & & & \vert \\
        \hline 
    \text{---} & a & \text{---} & 0
   \end{pNiceArray} = \]
   \[\det Y \det X + \left(\prod_i d_i\right) \det \begin{pNiceArray}{ccc|c}
        & & & \vert \\
        & X & & b     \\
        & & & \vert \\
        \hline 
    \text{---} & a & \text{---} & 0
   \end{pNiceArray}.\]
\end{proof}

\begin{corollary}
    If \(X\) is nilpotent, and
    \[\begin{pNiceArray}{ccc|ccc}
        & & & & & \vert \\
        & X & & & & b    \\
        & & & & & \vert \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y & \\
       & & &  \\
       \end{pNiceArray}\]
    is nilpotent  as well, then \(Y\) is nilpotent (TODO: and that other determinant is zero).
    \begin{proof}
        By the previous lemma, the characteristic polynomial of the big matrix is
        \[g_X(\lambda) g_Y(\lambda) + f(\lambda),\]
        where \(g_X(\lambda) = \lambda^m\) is the characteristic polynomial of \(X\), and \(f(\lambda)\) is some polynomial of degree at most \(m - 1\).
    \end{proof}
\end{corollary}
\par Now, taking the previous corolloary and the first lemma together, we see that 
\[\mathcal{S}_{m,n}' = \left\{\begin{pNiceArray}{ccc|cccc}
    & & & & & & \vert \\
    & X & & & & & b    \\
    & & & & & & \vert \\
   \hline
   \text{---} & a & \text{---} & 0 \\
    & & & 1 & 0  & \\
    &  & & & \ddots & \ddots  \\
    & & & & & 1 & 0
   \end{pNiceArray} : a, b \in \mathbb{C}^m, X \in \gl_m \textrm{ is nilpotent}\right\}.\]

@@

    Let \(P = \max_{i \in I} \rho_i\), and for \(i \in I\), let \(P_i = \max_{j \in I : j > i} \rho_j\).

@@

        \item The chain of length \(n + P\) beginning with \(f_n - (A^{n + P}f_n \rightshift n + P)\)
        \item For \(i \in I\), the chain of length \(q_i + P_i\) beginning with \(v_i - (A^{q_i + P_i} v_i \rightshift q_i + P_i)\), where \(v_i := A^{n - q_i} f_n - \sum_{l = 1}^{p_i} b_{il} e_{i,l + q_i}\)

@@

    \par We have defined \(P\) and \(P_i\) in terms of the \(\rho_i\) to make it clear that they only depend on \(I\) and the \(\rho_i\).
    Now we provide more useful characterizations.
    It is immediate from the definition of \(U_{I, (\rho_i)}\) that \(P = \max_{l : q_l < n} p_l\), and \(P_i = \max_{l : q_l < q_i} p_l\).

@@

        \par Now consider the chain beginning with \(f_n - (A^{n + P} f_n \rightshift n + P)\).
        Note that \(A^n f_n = b\), so \(A^{n + P} f_n = A^P b = b \leftshift P\).
        By shifting \(b\) left \(P\) times, we zero out all the rows \(i\) where \(q_i < n\).
        This ensures that the operation of shifting \(b \leftshift P\) right \(n + P\) times is invertible by shifting left \(n + P\) times.

@@

        \[A^{n + P} f_n = b \leftshift P = ((b \leftshift P) \rightshift n + P) \leftshift n + P = A^{n + P}(A^{n + P}f_n \rightshift n + P).\]
        This shows that the chain has length at most \(n + P\), as desired.
        \par Now, let \(i \in I\).
        We have \(v_i = A^{n - q_i} f_n - \sum_{l = 1}^{p_i} b_{il} e_{i,l + q_i}\).
        First we note that \(q_i < n\) by the definition of \(U_{I, (\rho_i)}\), so the definition of \(v_i\) makes sense.
        We consider the chain beginning with \(v_i - (A^{q_i + P_i} v_i \rightshift q_i + P_i)\).
        Note that \(A^{q_i} v_i\) is just \(b\) with row \(i\) zeroed out.
        For brevity, we write \(b_i := A^{q_i} v_i\).
        Note that \(b_i \leftshift P_i\) has rows \(l\) zeroed out, for all \(l\) with \(q_l < q_i\).
        This ensures that shifting \(b_i \leftshift P_i\) right \(q_i + P_i\) times can be inverted by shifting left \(q_i + P_i\) times.

@@

        \[A^{q_i + P_i} v_i = b_i \leftshift P_i = ((b_i \leftshift P_i) \rightshift q_i + P_i) \leftshift q_i + P_i = A^{q_i + P_i}(A^{q_i + P_i} v_i \rightshift q_i + P_i).\]
        This shows that the chain has length at most \(q_i + P_i\), as desired.

>>>>

\subsection{A determinant}
\begin{lemma}
    Let \(X \in \gl_m\), and let 
    \[Y = \begin{pmatrix}
        y_{11} & y_{12} & y_{13} & \cdots & y_{1,n-1} & y_{1n} \\
        d_1 & y_{22} & y_{23} & \cdots & y_{2,n-1} & y_{2n} \\
            & d_2   & y_{33} & \cdots & y_{3,n-1} & y_{3n} \\
            & & \ddots & \cdots & \vdots  & \vdots \\
            & & & d_{n - 2} & y_{n-1,n-1} & y_{n - 1,n}\\
            & & & & d_{n - 1} & y_{nn}
    \end{pmatrix} \in \gl_n.\]
    For any \(a, b \in \mathbb{C}^m\),
    \[\det 
    \begin{pNiceArray}{ccc|ccc}
        & & & & & \vert \\
        & X & & & & b    \\
        & & & & & \vert \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y & \\
       & & &  \\
       \end{pNiceArray} = \]
       \[\det X \det Y + \left(\prod_i d_i\right) \det \begin{pNiceArray}{ccc|c}
            & & & \vert \\
            & X & & b     \\
            & & & \vert \\
            \hline 
        \text{---} & a & \text{---} & 0
       \end{pNiceArray}\]
\end{lemma}
\begin{proof}
    By induction on \(n\).
    In the case \(n = 1\), expanding along the last row (taking the usual interpretation of the empty product) gives the desired result.
    \par Now suppose \(n > 1\).
    Expanding along the last row, we get 
    \[d_{n - 1} \det \begin{pNiceArray}{ccc|ccc}
        & & & & & \vert \\
        & X & & & & b    \\
        & & & & & \vert \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y_{n,n-1} & \\
       & & &  \\
       \end{pNiceArray} -
    y_{nn} \det \begin{pNiceArray}{ccc|ccc}
        & & & & & \\
        & X & & & &   \\
        & & & & & \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y_{n,n} & \\
       & & &  \\
       \end{pNiceArray}.\]
    Using our inductive hypothesis for the first determiniant, and using that \(\det \begin{pNiceArray}{c|c}
        A_{11} & 0 \\
        A_{21} & A_{22}
    \end{pNiceArray} = \det A_{11} \det A_{22}\) for the second, the expression becomes
    \[d_{n - 1} \left(\det X \det Y_{n,n-1} + \left(\prod_{i \leq n - 2} d_i\right) \det \begin{pNiceArray}{ccc|c}
        & & & \vert \\
        & X & & b     \\
        & & & \vert \\
        \hline 
    \text{---} & a & \text{---} & 0
   \end{pNiceArray}\right)
   - y_{nn} \det X \det Y_{nn} = \]
   \[(d_{n - 1} Y_{n,n-1} - y_{nn} \det Y_{nn})\det X + \left(\prod_i d_i\right) \det \begin{pNiceArray}{ccc|c}
        & & & \vert \\
        & X & & b     \\
        & & & \vert \\
        \hline 
    \text{---} & a & \text{---} & 0
   \end{pNiceArray} = \]
   \[\det Y \det X + \left(\prod_i d_i\right) \det \begin{pNiceArray}{ccc|c}
        & & & \vert \\
        & X & & b     \\
        & & & \vert \\
        \hline 
    \text{---} & a & \text{---} & 0
   \end{pNiceArray}.\]
\end{proof}

\begin{corollary}
    If \(X\) is nilpotent, and
    \[\begin{pNiceArray}{ccc|ccc}
        & & & & & \vert \\
        & X & & & & b    \\
        & & & & & \vert \\
       \hline
       \text{---} & a & \text{---} &  \\
       & & & & Y & \\
       & & &  \\
       \end{pNiceArray}\]
    is nilpotent  as well, then \(Y\) is nilpotent.
    If all the \(d_i\) are nonzero, then
    \[\begin{pNiceArray}{ccc|c}
        & & & \vert \\
        & X & & b     \\
        & & & \vert \\
        \hline 
    \text{---} & a & \text{---} & 0
   \end{pNiceArray}\]
   is nilpotent as well.
    \begin{proof}
        By the previous lemma, the characteristic polynomial of the big matrix is
        \[g_X(\lambda) g_Y(\lambda) + f(\lambda),\]
        where \(g_X(\lambda) = \lambda^m\) is the characteristic polynomial of \(X\), and \(f(\lambda)\) is some polynomial of degree at most \(m - 1\).
    \end{proof}
\end{corollary}

@@

\par Now, taking the previous corolloary and the first lemma together, we see that 
\[\mathcal{S}_{m,n}' = \left\{\begin{pNiceArray}{ccc|cccc}
    & & & & & & \vert \\
    & X & & & & & b    \\
    & & & & & & \vert \\
   \hline
   \text{---} & a & \text{---} & 0 \\
    & & & 1 & 0  & \\
    &  & & & \ddots & \ddots  \\
    & & & & & 1 & 0
   \end{pNiceArray} : a, b \in \mathbb{C}^m, X \in \gl_m \textrm{ is nilpotent}\right\}.\]

@@

@@

        \item The chain of length \(n + p_{i_1}\) beginning with \(f_n - (A^{n + p_{i_1}}f_n \rightshift n + p_{i_1})\)
        \item For \(k \in \{1, ..., r\}\), the chain of length \(q_{i_k} + p_{i_{k + 1}}\) beginning with \(v_{i_k} - (A^{q_{i_k} + p_{i_{k + 1}}} v_{i_k} \rightshift q_{i_k} + p_{i_{k + 1}})\), where \(v_{i_k} := A^{n - q_{i_k}} f_n - \sum_{l = 1}^{p_{i_k}} b_{i_kl} e_{i,l + q_{i_k}}\)

@@

@@

        \par Now consider the chain beginning with \(f_n - (A^{n + p_{i_1}} f_n \rightshift n + p_{i_1})\).
        Note that \(A^n f_n = b\), so \(A^{n + p_{i_1}} f_n = A^{p_{i_1}} b = b \leftshift p_{i_1}\).
        By shifting \(b\) left \(p_{i_1}\) times, we zero out all the rows \(i\) where \(q_i < n\).
        This ensures that the operation of shifting \(b \leftshift p_{i_1}\) right \(n + p_{i_1}\) times is invertible by shifting left \(n + p_{i_1}\) times.

@@

            A^{n + p_{i_1}} f_n & = \\
            b \leftshift p_{i_1} & = \\
            ((b \leftshift p_{i_1}) \rightshift n + p_{i_1}) \leftshift n + p_{i_1} & = \\
            A^{n + p_{i_1}}(A^{n + p_{i_1}}f_n \rightshift n + p_{i_1}) & .
        This shows that the chain has length at most \(n + p_{i_1}\), as desired.
        \par Now, let \(k \in \{1, ..., r\}\).
        We have \(v_{i_k} = A^{n - q_{i_k}} f_n - \sum_{l = 1}^{p_{i_k}} b_{i_kl} e_{i_k,l + q_{i_k}}\).
        First we note that \(q_{i_k} < n\) by the definition of \(U_{I, (\rho_i)}\), so the definition of \(v_{i_k}\) makes sense.
        We consider the chain beginning with \(v_{i_k} - (A^{q_{i_k} + p_{i_{k + 1}}} v_{i_k} \rightshift q_{i_k} + p_{i_{k + 1}})\).
        Note that \(A^{q_{i_k}} v_{i_k}\) is just \(b\) with row \(i_k\) zeroed out.
        For brevity, we write \(b_{i_k} := A^{q_{i_k}} v_{i_k}\).
        Note that \(b_{i_k} \leftshift p_{i_{k + 1}}\) has rows \(l\) zeroed out, for all \(l\) with \(q_l < q_{i_k}\).
        This ensures that shifting \(b_{i_k} \leftshift p_{i_{k + 1}}\) right \(q_{i_k} + p_{i_{k + 1}}\) times can be inverted by shifting left \(q_{i_k} + p_{i_{k + 1}}\) times.

@@

            A^{q_{i_k} + p_{i_{k + 1}}} v_{i_k} & = \\
            b_{i_k} \leftshift p_{i_{k + 1}} & = \\
            ((b_{i_k} \leftshift p_{i_{k + 1}}) \rightshift q_{i_k} + p_{i_{k + 1}}) \leftshift q_{i_k} + p_{i_{k + 1}} & = \\
            A^{q_{i_k} + p_{i_{k + 1}}}(A^{q_{i_k} + p_{i_{k + 1}}} v_{i_k} \rightshift q_{i_k} + p_{i_{k + 1}}) & .
        This shows that the chain has length at most \(q_{i_k} + p_{i_{k + 1}}\), as desired.
